Aneuploidy is an evolutionary sidetrack on the path to adaptation

### 2022-07-18 ms/ms.tex
start reorganize, add mutation results

% NatBib
\usepackage[round,colon,authoryear]{natbib}

% Title page
% Chromosomal duplication is a transient evolutionary solution to stress
%\title{Clonal interference between aneuploidy and mutation}
\title{Adaptive evolution with aneuploidy and mutation}

@@

\paragraph*{Aneuploidy is common in eukaryotes.}
Aneuploidy is an imbalance in the number of chromosomes in the cell: an incorrect karyotype.
Evidence suggests aneuploidy is very common in eukaryotes, e.g. animals~\citep{Santaguida2015review, Naylor2016, Bakhoum2017}, and fungi~\citep{Pavelka2010, Zhu2016, Robbins2017, Todd2017}.
Aneuploidy has been implicated in cancer formation and progression~\citep{Boveri2008, Schvartzman2010}: 
90\% of solid tumours and 50\% of blood cancers are aneuploid~\citep{Santaguida2015review}.
Aneuploidy is also linked to the emergence of drug resistance~\citep{Selmecki2009} and virulence~\citep{Moller2018} in fungal pathogens, which are under-studied~\citep{Rodrigues2018} despite infecting close to a billion people per year, causing serious infections and significant morbidity in >150 million people per year and killing >1.5 million people per year~\citep{Selmecki2009, Rodrigues2018}.
In addition, aneuploidy is common in protozoan pathogens of the Leishmania genus, a major global health concern~\citep{Mannaert2012}.

\paragraph*{Aneuploidy is generally deleterious.}
The molecular and genetic mechanisms involved in aneuploidy have been explored~\citep{Musacchio2007, Sheltzer2011, Chen2012b, Rancati2013, Gerstein2015, Shor2015}.
Experiments with human and mouse embryos found that aneuploidy is usually lethal.
It is also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011}. For example, aneuploid mouse embryonic cells grow slower than euploid cells~\citep{Williams2008}.
Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall fitness~\citep{Niwa2006, Torres2007, Pavelka2010, Sheltzer2011, Kasuga2016}, in part due to proteotoxic stress caused by increased expression in aneuploid cells~\citep{Pavelka2010, Santaguida2015, Zhu2018} and hypo-osmotic-like stress~\citep{Tsai2019}.
% TODO cite Yang, F. et al. The fitness costs and benefits of trisomy of each Candida albicans chromosome. Genetics 218, 1–7 (2021).

\paragraph*{Aneuploidy can lead to adaptation.}
However, aneuploidy can be beneficial under stressful conditions due to the wide range of phenotypes it can produce, some of which are advantageous~\citep{Pavelka2010}.
Thus, aneuploidy can lead to rapid adaptation in unicellular eukaryotes~\citep{Gerstein2015,Torres2010, Hong2014, Rancati2008}, as well as to rapid growth of somatic tumour cells~\citep{Schvartzman2010, Sheltzer2017}.
For example, aneuploidy in \yeast facilitates adaptation to a variety of stressful conditions like heat and pH~\citep{Yona2012}, copper~\citep{Covo2014, Gerstein2015}, salt~\citep{Dhar2011}, and nutrient limitation~\citep{Dunham2002, Gresham2008}.
Importantly, aneuploidy can also lead to drug resistance in pathogenic fungi such as \calbicans~\citep{Selmecki2008, Selmecki2010, Gerstein2018} and \cneoformans~\citep{Sionov2010}, which cause candidiasis and meningoencephalitis, respectively.

\paragraph*{Transient adaptive solution.} 
Aneuploidy differs from mutation due to its distinct properties. 
Chromosome duplication usually occurs more often than mutation and on average produces larger fitness effects.
Yet, because it affects many genes on a whole chromosome or a chromosome fragment, aneuploidy also carries fitness costs.
Thus, aneuploidy can be a \emph{transient adaptive solution}: it can rapidly occur and fix in the population under stressful conditions, and can be rapidly lost when the cost outweighs the benefit---when stress is removed or after beneficial mutations occur.
Experimental evidence of such a transient role of aneuploidy was demonstrated by \citet{Yona2012}. They evolved populations of \yeast under strong heat or pH stress.
The populations adapted to the heat and pH stress within 450 and 150 generations, and this adaptation was determined to be due to chromosome duplications.
Much later, after more than 1500 and 750 generations, for the heat and pH stress, respectively, the populations reverted back to an euploid state, while remaining adapted to the stress and accumulating multiple mutations.
However, under gradual heat stress, aneuploidy was not observed.
\citet{Yona2012} concluded that aneuploidy serves as a transient adaptive solution, or a ``quick fix'', which is expected to facilitate adaptation.

\paragraph*{The present study.}
Here, we develop an evolutionary-genetic model that includes the effects of natural selection, genetic drift, aneuploidy, and mutation to examine the role of aneuploidy in adaptive evolution.
This model follows a population of cells characterised by both their ploidy and their genotype.
We fit this model to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including rates and fitness effects of aneuploidy and mutation, and to test hypotheses about the evolutionary process by comparing different versions of the model and predicting the genotype frequency dynamics using the estimated parameter values.

@@

@@

%\pagebreak
% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, Uri Obolski, and Tal Simon for discussions and comments.
This work was supported in part by 
the Israel Science Foundation (YR 552/19),
Minerva Stiftung Center for Lab Evolution (YR),
Minerva Stiftung short-term research grant (MP).
% TODO funding for Avihu
}

@@

>>>>

% NatBib
\usepackage[super,comma]{natbib}
%\usepackage[round,colon,authoryear]{natbib}

% Title page
% Chromosomal duplication is a transient evolutionary solution to stress
\title{Aneuploidy is an evolutionary sidetrack on the path to adaptation}

@@

%\paragraph*{Aneuploidy is common in eukaryotes.}
Aneuploidy is an imbalance in the number of chromosomes in the cell: an incorrect karyotype.
Evidence suggests aneuploidy is very common in eukaryotes, e.g. animals~\citep{Santaguida2015review, Naylor2016, Bakhoum2017}, and fungi~\citep{Pavelka2010, Zhu2016, Robbins2017, Todd2017}.
Aneuploidy has been implicated in cancer formation and progression~\citep{Boveri2008, Schvartzman2010,Santaguida2015review}.
%90\% of solid tumours and 50\% of blood cancers are aneuploid~\citep{Santaguida2015review}.
It is also common in protozoan pathogens of the Leishmania genus, a major global health concern~\citep{Mannaert2012}, and contributes to the emergence of drug resistance~\citep{Selmecki2009} and virulence~\citep{Moller2018} in fungal pathogens, which are under-studied~\citep{Rodrigues2018}, despite infecting a billion people per year, causing significant morbidity in >150 million and death in >1.5 million people per year~\citep{Selmecki2009, Rodrigues2018}.

%\paragraph*{Aneuploidy is generally deleterious.}
%The molecular and genetic mechanisms involved in aneuploidy have been explored~\citep{Musacchio2007, Sheltzer2011, Chen2012b, Rancati2013, Gerstein2015, Shor2015}.
Experiments with human and mouse embryos found that aneuploidy is usually lethal.
It is also associated with developmental defects and lethality in other multicellular organisms~\citep{Sheltzer2011}. For example, aneuploid mouse embryonic cells grow slower than euploid cells~\citep{Williams2008}.
Similarly, in unicellular eukaryotes growing in benign conditions, aneuploidy usually leads to slower growth and decreased overall fitness~\citep{Niwa2006, Torres2007, Pavelka2010, Sheltzer2011, Kasuga2016, Yang2021}, in part due to proteotoxic stress caused by increased expression in aneuploid cells~\citep{Pavelka2010, Santaguida2015, Zhu2018} and hypo-osmotic-like stress~\citep{Tsai2019}.

%\paragraph*{Aneuploidy can lead to adaptation.}
However, aneuploidy can be beneficial under stressful conditions due to the wide range of phenotypes it can produce, some of which are advantageous~\citep{Pavelka2010, Yang2021}.
Thus, aneuploidy can lead to rapid adaptation in unicellular eukaryotes~\citep{Gerstein2015,Torres2010, Hong2014, Rancati2008}, as well as to rapid growth of somatic tumour cells~\citep{Schvartzman2010, Sheltzer2017}.
For example, aneuploidy in \yeast facilitates adaptation to a variety of stressful conditions like heat and pH~\citep{Yona2012}, copper~\citep{Covo2014, Gerstein2015}, salt~\citep{Dhar2011}, and nutrient limitation~\citep{Dunham2002, Gresham2008, Avecilla2022}, with similar results in \calbicans~\citep{Yang2021}.
Importantly, aneuploidy can also lead to drug resistance in pathogenic fungi such as \calbicans~\citep{Selmecki2008, Selmecki2010, Gerstein2018} and \cneoformans~\citep{Sionov2010}, which cause candidiasis and meningoencephalitis, respectively.

%\paragraph*{Transient adaptive solution.} 
Experimental evidence for the role of aneuploidy in adaptive evolution was demonstrated by \citet{Yona2012}. They evolved populations of \yeast under strong heat stress.
The populations adapted to the heat stress within 450 generations, and this adaptation was determined to be due a duplication of chromosome III.
Later on, after more than 1,500 generations, the populations reverted back to an euploid state, while remaining adapted to the heat stress. 
The authors hypothesized that aneuploidy is a \emph{transient adaptive solution}, because it can rapidly appear and fixate in the population under stressful conditions, and can then be rapidly lost when the cost of aneuploidy outweighs its benefit---after the stress is removed, or after "refined" beneficial mutations appear and fixate.

%\paragraph*{The present study.}
Here, we test the hypothesis that aneuploidy is a \emph{transient adaptive solution}.
First, we analyze previously unpublished sequencing data from the original experimental populations of \citet{Yona2012} to assess if the aneuploid population and the evolved euploid population carry the same mutant alleles, as expected if the later is descended from the former.
Second, we develop an evolutionary genetic model and fit it to the experimental results of \citet{Yona2012} in order to predict the genotype frequency dynamics in the experimental populations, thereby estimating the frequency of evolved euploid cells that descended from aneuploid cells. 
Our results show that, despite aneuploidy having reached high frequencies in the experimental populations, the majority of cells in the evolved euploid populations likely did not descend from aneuploid cells, but rather directly from wild type euploid cells.
These results suggest that aneuploidy may have been a sidetrack, rather than a stepping stone, on the path to adaptation.

\paragraph{DNA sequencing.}

In the heat-stress experiment of \citet{Yona2012}, four populations of \yeast evolved under \SI{39}{\celsius}. Aneuploidy fixed in all four experimental repetitions in the first 450 generations. Two of the repetitions, marked $H2$ and $H4$, carried no large-scale duplications other than a chromosome III trisomy. These two repetitions continued to evolve under the same conditions, wherein aneuploidy was eliminated by generation 1,700 and 2,350.
For each of these two populations we sequenced the ancestral population (generation zero), the aneuploid population (generation 450), and the evolved euploid population (generation 1,700 or 2,350) to estimate the mutant allele frequencies (\Cref{table:mutations}).
Overall, between 100 and 173 mutant alleles were detected with at least a single read in the six population that were sampled. 
Disregarding 45 and 40 alleles that were present in the ancestral populations at a frequency >10\%, the aneuploid and euploid populations carried a large number of mutant alleles: 82 and 95, respectively, in repetition $H2$, and 60 and 66 in repetition $H4$.

Surprisingly, out of all these mutant alleles, none was present at a frequency >20\% in both the aneuploid and the evolved euploid populations. 
Moreover, a high mutant allele frequency in the aneuploid population implies a low frequency in the evolved euploid population, and vice-versa (Spearman's correlation coefficient $\rho=-0.64$ and $-0.66$ in the two experimental repetitions; \Cref{fig:mutation_frequencies}). 
Specifically, for the 18 mutant alleles with high frequency in the aneuploid populations (>20\%), the highest frequencies in the euploid populations were 15.4\%, 16\%, 16.3\% and 19.6\% (the rest were below 15\%).
Similarly, for the 48 mutant alleles with high frequency in the evolved euploid populations, the highest frequencies in the aneuploid populations were 2.7\%, 7.7\%, and 11.1\% (the rest were below 1\%). 
Thus, mutant alleles frequent in the aneuploid populations decreased in frequency when aneuploidy was lost, and most evolved euploid cells are likely not descendent from aneuploid~cells.

%%%%%%%% Fig mutation frequencies
\begin{figure}[h!]
  \centering
    \includegraphics[width=\textwidth]{../figures/mutation_frequencies.pdf}      
  \caption{
    \textbf{Frequencies of mutant alleles in the experimental populations are negatively correlated.} 
    Frequencies of mutant alleles when trisomy was widespread in the population (y-axis) and after it was eliminated (x-axis) in two experimental repetitions (circles for $H2$ and triangles for $H4$) from \citet{Yona2012}. Mutant alleles with >20\% in the aneuploid population were <20\% in the euploid population, and vice versa (the upper-right quadrant is empty), suggesting that the majority of evolved euploid cells did not descend from the most common aneuploid genotypes. 
Alleles with frequency below and above 10\% in the ancestral populations are in black and gray, respectively. 
Solid black line is a linear orthogonal distance regression line (slope=$-0.559$, intercept=$0.164$).
Dashed vertical and horizontal lines show allele frequencies 20\%. Error bars show standard error of the mean accounting for the number of reads.
}
  \label{fig:mutation_frequencies}
\end{figure}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\paragraph{Evolutionary genetic model.} 
% TODO continue here

We develop an evolutionary genetic model that that includes the effects of natural selection, genetic drift, aneuploidy, and mutation to examine the role of aneuploidy in adaptive evolution.
The model follows a population of cells characterised by both their ploidy and their genotype, fit this model to the experimental results of \citet{Yona2012} using an \emph{approximate Bayesian computation} framework~\citep{Sisson2009} to infer model parameters, including rates and fitness effects of aneuploidy and mutation, and to test hypotheses about the evolutionary process by comparing different versions of the model and predicting the genotype frequency dynamics using the estimated parameter values.

@@

% quote: This scenario may thus imply that the duplication of a cer- tain chromosome may provide the evolutionary time window needed for the population to search for solutions not only on the duplicated chromosome but throughout the genome.

\citet{Yona2012} have suggested that the aneuploid cells that come to dominate the experimental populations were later replaced by ``refined strains, which eliminated the cost associated with the trisomy, exhibit improved growth both at the permissive and the high temperatures compared with their trisomic ancestors'' and that ``it is expected that the original contribution of trisomy to heat tolerance was replaced by alternative solutions.''. 
However, our results suggest that most of the cells in the evolved euploid population do not carry mutations that were common in the aneuploid populations, but rather a different set of mutations.
Therefore, the majority of the evolved euploid cells are not descended from aneuploid cells, but rather are direct descendants of the ancestral population.

\cite{Yona2012} have analyzed the expression levels of 14 heat-shock proteins (HSP). They found that their expression level changed ``very modestly'' in the aneuploid populations, but that most of them ``showed significant up-regulation'' in the evolved euploid population.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% Acknowledgements
{\small
\section*{Acknowledgements}
We thank Yitzhak Pilpel, Orna Dahan, Lilach Hadany, Judith Berman, David Gresham, Shay Covo, Martin Kupiec, Uri Obolski, and Tal Simon for discussions and comments.
This work was supported in part by 
the Israel Science Foundation (ISF, YR 552/19),
the US--Israel Binational Science Foundation (BSF, YR 2021276)
Minerva Stiftung center for Lab Evolution (YR),
Minerva Stiftung short-term research grant (MP).
}

@@

%TODO table:mutations
